---
title: "Integration with cards/cardx ARD Framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration with cards/cardx ARD Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The **beeca** package produces Analysis Results Data (ARD) in a CDISC-inspired format through the `marginal_results` component. This vignette explores the relationship between beeca's ARD structure and the more general ARD framework provided by the [insightsengineering/cards](https://github.com/insightsengineering/cards) and [insightsengineering/cardx](https://github.com/insightsengineering/cardx) packages.

## ARD Frameworks Comparison

### beeca ARD Structure

The beeca package creates ARDs following pharmaceutical industry conventions for clinical trial reporting:

```{r setup, eval=FALSE}
library(beeca)

trial01$trtp <- factor(trial01$trtp)
fit1 <- glm(aval ~ trtp + bl_cov, family = "binomial", data = trial01) |>
  get_marginal_effect(trt = "trtp", method = "Ye", contrast = "diff", reference = "0")

fit1$marginal_results
```

**beeca ARD Schema:**

| Column | Type | Description |
|--------|------|-------------|
| `TRTVAR` | character | Treatment variable name |
| `TRTVAL` | character | Treatment level value |
| `PARAM` | character | Parameter/outcome variable name |
| `ANALTYP1` | character | Analysis type (DESCRIPTIVE or INFERENTIAL) |
| `STAT` | character | Statistic identifier (N, n, %, risk, risk_se, diff, diff_se) |
| `STATVAL` | numeric | Numeric value of the statistic |
| `ANALMETH` | character | Analysis method (count, percentage, g-computation, HC0, etc.) |
| `ANALDESC` | character | Analysis description including package version |

**Key Features:**
- CDISC ADaM-inspired column names (PARAM, ANALTYP1)
- Compact structure optimized for clinical trial reporting
- Single value per row (atomic numeric type)
- Direct integration with CDISC workflows

### cards/cardx ARD Structure

The cards package provides a more general ARD framework for statistical computing:

**cards ARD Schema:**

| Column | Description |
|--------|-------------|
| `group1` | First grouping variable name |
| `group1_level` | Level/value of first grouping variable |
| `group2` | Second grouping variable name (optional) |
| `group2_level` | Level/value of second grouping variable (optional) |
| `variable` | The analysis variable name |
| `variable_level` | Specific level of the variable (for categorical) |
| `stat_name` | Statistical measure identifier |
| `stat_label` | Human-readable label for the statistic |
| `stat` | The calculated value (list-column) |
| `context` | Analysis context metadata |
| `fmt_fn` | Formatting function reference (list-column) |
| `warning` | Warnings from calculation (list-column) |
| `error` | Errors from calculation (list-column) |

**Key Features:**
- List-column structure for flexible data types
- Built-in error/warning capture
- Human-readable labels separate from internal names
- Formatting functions attached to each statistic
- Extensible context system

## Mapping Between Frameworks

| beeca | cards | Notes |
|-------|-------|-------|
| `TRTVAR` | `group1` | Treatment variable name |
| `TRTVAL` | `group1_level` | Treatment level |
| `PARAM` | `variable` | Outcome variable |
| `STAT` | `stat_name` | Statistic identifier |
| `STATVAL` | `stat` | Value (numeric vs list-column) |
| `ANALTYP1` + `ANALMETH` | `context` | Analysis context |
| — | `stat_label` | Missing in beeca |
| — | `fmt_fn` | Missing in beeca |
| — | `warning`, `error` | Missing in beeca |
| `ANALDESC` | attribute | Package metadata |

## Converting beeca ARD to cards Format

For users who want to work with both frameworks, here's a conversion function:

```{r conversion-function, eval=FALSE}
#' Convert beeca marginal_results to cards ARD format
#'
#' @param marginal_results A beeca marginal_results tibble
#' @return A cards ARD object
#' @export
beeca_to_cards_ard <- function(marginal_results) {

  if (!requireNamespace("cards", quietly = TRUE)) {
    stop("Package 'cards' is required for this conversion. Please install it.")
  }

  # Create stat_label mapping
  stat_label_map <- c(
    "N" = "N",
    "n" = "n",
    "%" = "%",
    "risk" = "Risk",
    "risk_se" = "Risk SE",
    "diff" = "Difference",
    "diff_se" = "Difference SE",
    "rr" = "Risk Ratio",
    "rr_se" = "Risk Ratio SE",
    "or" = "Odds Ratio",
    "or_se" = "Odds Ratio SE",
    "logrr" = "Log Risk Ratio",
    "logrr_se" = "Log Risk Ratio SE",
    "logor" = "Log Odds Ratio",
    "logor_se" = "Log Odds Ratio SE"
  )

  result <- marginal_results |>
    dplyr::rename(
      group1 = TRTVAR,
      group1_level = TRTVAL,
      variable = PARAM,
      stat_name = STAT
    ) |>
    dplyr::mutate(
      # Add human-readable labels
      stat_label = dplyr::case_when(
        stat_name %in% names(stat_label_map) ~ stat_label_map[stat_name],
        TRUE ~ toupper(stat_name)
      ),

      # Consolidate context from multiple beeca columns
      context = paste(
        tolower(ANALTYP1),
        ANALMETH,
        sep = "_"
      ),

      # Convert atomic numeric to list column (required by cards)
      stat = as.list(STATVAL),

      # Add empty list columns for cards compatibility
      fmt_fn = list(NULL),
      warning = list(NULL),
      error = list(NULL),

      # Add variable_level (not used in beeca's current structure)
      variable_level = NA_character_
    ) |>
    dplyr::select(
      group1, group1_level, variable, variable_level,
      stat_name, stat_label, stat, context,
      fmt_fn, warning, error
    )

  # Store beeca metadata as attributes
  attr(result, "beeca_description") <- unique(marginal_results$ANALDESC)

  # Apply cards column ordering and class
  result |>
    cards::tidy_ard_column_order() |>
    cards::as_card()
}
```

## Example Usage

```{r example-usage, eval=FALSE}
# Fit model and get beeca results
trial01$trtp <- factor(trial01$trtp)
fit1 <- glm(aval ~ trtp + bl_cov, family = "binomial", data = trial01) |>
  get_marginal_effect(trt = "trtp", method = "Ye", contrast = "diff", reference = "0")

# Convert to cards format
cards_ard <- beeca_to_cards_ard(fit1$marginal_results)

# Now can use cards utilities
cards::print_ard(cards_ard)

# Bind with other cards ARDs
combined_ard <- cards::bind_ard(
  cards_ard,
  cards::ard_continuous(trial01, by = trtp, variables = bl_cov)
)
```

## Use Cases for Integration

### 1. Comprehensive Reporting

Combine beeca's covariate-adjusted treatment effects with descriptive statistics from cards:

```{r comprehensive-reporting, eval=FALSE}
# Treatment effect from beeca
te_ard <- beeca_to_cards_ard(fit1$marginal_results)

# Baseline characteristics from cards
baseline_ard <- cards::ard_continuous(
  data = trial01,
  by = trtp,
  variables = c(bl_cov, age, weight)
)

# Combine into comprehensive ARD
full_ard <- cards::bind_ard(te_ard, baseline_ard)
```

### 2. Quality Control Workflows

Use cards error-handling features with beeca results:

```{r qc-workflow, eval=FALSE}
# Multiple analyses with error capture
analyses <- lapply(contrasts, function(contrast_type) {
  cards::eval_capture_conditions({
    glm(aval ~ trtp + bl_cov, family = "binomial", data = trial01) |>
      get_marginal_effect(trt = "trtp", contrast = contrast_type) |>
      (\(x) x$marginal_results)()
  })
})

# Check for errors
errors <- Filter(function(x) !is.null(x$error), analyses)
```

### 3. Cross-Study Meta-Analysis

Standardize results across multiple studies:

```{r meta-analysis, eval=FALSE}
# Study 1 (beeca)
study1_ard <- beeca_to_cards_ard(fit_study1$marginal_results) |>
  dplyr::mutate(study = "STUDY001")

# Study 2 (beeca)
study2_ard <- beeca_to_cards_ard(fit_study2$marginal_results) |>
  dplyr::mutate(study = "STUDY002")

# Combine
meta_ard <- cards::bind_ard(study1_ard, study2_ard)

# Extract estimates for meta-analysis
estimates <- meta_ard |>
  dplyr::filter(stat_name == "diff") |>
  dplyr::mutate(estimate = unlist(stat))
```

## When to Use Each Format

### Use beeca's Native ARD When:
- Working primarily in CDISC environments
- Generating clinical trial reports
- Compact storage is preferred
- Integrating with ADaM datasets
- Direct numeric values are sufficient

### Convert to cards ARD When:
- Combining with other cards/cardx analyses
- Need error/warning tracking
- Want formatting functions attached
- Working with general statistical workflows
- Need extensible metadata system
- Generating complex multi-analysis reports

## Design Philosophy Differences

### beeca ARD
- **Goal:** Clinical trial reporting efficiency
- **Structure:** Compact, CDISC-aligned
- **Storage:** Atomic numeric values
- **Context:** Domain-specific columns (ANALTYP1, ANALMETH)
- **Focus:** Marginal treatment effects

### cards ARD
- **Goal:** General statistical computing reproducibility
- **Structure:** Flexible, extensible
- **Storage:** List-columns for any data type
- **Context:** Single extensible context column
- **Focus:** Any statistical analysis

## Future Enhancements

Potential additions to beeca for better cards integration:

1. **S3 Method:** `as_card.beeca()` for seamless conversion
2. **Native cards Output:** Optional `output = "cards"` parameter in `get_marginal_effect()`
3. **Error Capture:** Wrap internal calculations with `eval_capture_conditions()`
4. **Formatting Functions:** Attach default formatting to each statistic type
5. **Stat Labels:** Add human-readable labels to ARD by default

## References

- [cards package documentation](https://insightsengineering.github.io/cards/)
- [cardx package documentation](https://insightsengineering.github.io/cardx/)
- [CDISC ADaM Implementation Guide](https://www.cdisc.org/standards/foundational/adam)
- [beeca package documentation](https://openpharma.github.io/beeca/)

## Session Info

```{r session-info}
sessionInfo()
```
